Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-28 paper.qmd
Update paper.qmd

Other [R]{.proglang} packages such as
[groundhog]{.pkg} [@simonsohn2023] and [rang]{.pkg} [@chan2023] enable
date-based package installation, while the Posit Package Manager provides dated
CRAN snapshots. However, like [renv]{.pkg}, these focus on R packages and do not
address system-level dependencies or R version management. 
Furthermore, using pre-compiled binaries of packages simplifies installation but do not eliminate the need for
system libraries required at runtime.

These tools represent significant progress in reproducibility within their
ecosystems, but they share a common limitation: none handles system-level
dependencies or the increasingly polyglot nature of modern data science. 
The era of a single tool dominating an entire analytical pipeline is giving 
way to a more pragmatic approach where researchers combine the complementary 
strengths of multiple languages [@peng2018]. This trend is particularly evident
in the synergistic use of [R]{.proglang}, with its rich ecosystem for statistical modeling, 
[Python]{.proglang}, with its extensive libraries for machine learning, and [Julia]{.proglang} 
for complex simulations through the use of interoperability libraries such as [PythonCall]{.pkg} 
(calling [Python]{.proglang} from [Julia]{.proglang}), [JuliaCall]{.pkg} 
(calling [Julia]{.proglang} from [Python]{.proglang} or [R]{.proglang}), [reticulate]{.pkg} 
(calling [Python]{.proglang} from [R]{.proglang}), [rpy2]{.pkg} and [ryp]{.pkg} (calling [R]{.proglang} from [Python]{.proglang}) [@osborne2024, @navarro2024].

Recent survey data confirms this trend, showing that research scientists use, on average, 
nearly two programming languages in their work, with the [R]{.proglang} and [Python]{.proglang} combination being 
the most prevalent [@chen2025]. This creates a significant challenge for reproducibility:
a project using [R]{.proglang}, [Python]{.proglang}, and system tools like
[Quarto]{.pkg} or \LaTeX requires coordinating multiple, disparate package managers, 
each with its own configuration and potential for conflict and using interoperability
libraries.

The demand for solutions that can manage this complexity is not merely academic; 
it is reflected in major strategic shifts within the industry. A prominent example is the 
evolution of Posit (formerly RStudio), which has pivoted from its origins as a premier R-centric 
development environment to a multi-language scientific data platform that treats Python as a 
first-class citizen. This industry trend, coupled with the proliferation of academic research
into polyglot tooling, underscores a critical reality: the central challenge for modern reproducible
science is no longer *if* polyglot workflows should be used, but *how* they can be managed in a unified, 
efficient, and reproducible manner.

>>>>

Other [R]{.proglang} packages such as
[groundhog]{.pkg} [@simonsohn2023] and [rang]{.pkg} [@chan2023] enable
date-based package installation, while the Posit Package Manager provides dated
CRAN snapshots. However, like [renv]{.pkg}, these focus on R packages and do not
address system-level dependencies or R version management. 
Furthermore, using pre-compiled binaries of packages simplifies installation but do not eliminate the need for
system libraries required at runtime. These tools represent significant progress in reproducibility within their
ecosystems, but they share a common limitation: none handles system-level
dependencies or the increasingly polyglot nature of modern data science.

Indeed, there is evidence that the era of a single tool dominating an entire analytical pipeline is giving 
way to a more pragmatic approach where researchers combine the complementary 
strengths of multiple languages [@peng2018]. This trend is particularly evident
in the synergistic use of [R]{.proglang}, with its rich ecosystem for statistical modeling, 
[Python]{.proglang}, with its extensive libraries for machine learning, and [Julia]{.proglang} 
for complex simulations through the use of interoperability libraries such as [PythonCall]{.pkg} 
(calling [Python]{.proglang} from [Julia]{.proglang}), [JuliaCall]{.pkg} 
(calling [Julia]{.proglang} from [Python]{.proglang} or [R]{.proglang}), [reticulate]{.pkg} 
(calling [Python]{.proglang} from [R]{.proglang}), [rpy2]{.pkg} and [ryp]{.pkg} (calling [R]{.proglang} from [Python]{.proglang}) [@osborne2024, @navarro2024].
Recent survey data confirms this trend, showing that research scientists use, on average, 
nearly two programming languages in their work, with the [R]{.proglang} and [Python]{.proglang} combination being 
the most prevalent [@chen2025]. This creates a significant challenge for reproducibility:
a project using [R]{.proglang}, [Python]{.proglang}, and system tools like
[Quarto]{.pkg} or \LaTeX requires coordinating multiple, disparate package managers, 
each with its own configuration and potential for conflict and using interoperability
libraries.

The demand for solutions that can manage this complexity is not merely academic; 
it is reflected in major strategic shifts within the industry. A prominent example is the 
evolution of Posit (formerly RStudio), which has pivoted from its origins as a premier R-centric 
service and software supplier to a company that treats [Python]{.proglang} as a 
first-class citizen alongside [R]{.proglang}. This industry trend, coupled with the proliferation of academic research
into polyglot tooling, underscores a critical reality: the central challenge for modern reproducible
science is no longer *if* polyglot workflows should be used, but *how* they can be managed in a unified, 
efficient, and reproducible manner.
